A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 MS.tex
Updates to make cross-referencing possible

@@

MRR estimates the length of time a mosquito remains alive and is still
in the area available for recapture. It thus provides a lower bound to
lifespan which we shall refer to as LBL. In 187 of the 230 MRR time
series the estimated LBL was less than 10 days (Fig. 1). The smallest
estimate was \textless{}1 day for the Asian malaria vector
\emph{Anopheles subpictus s.l.} which is unfeasibly short and almost
certainly reflects dispersal out of the recapture zone or a violation of
the assumptions of our analyses. The longest estimate was 18.3 days for
the temperate species \emph{Aedes simpsoni s.l.} which is a vector of
yellow fever in Africa. There are multiple data sets for the most
important vector species such as \emph{Anopheles gambiae, Aedes aegypti}
and \emph{albopictus} and \emph{Culex tarsalis} all of which show
considerable variation. For example, there are 54 estimates of LBL for
\emph{A. aegypti} which range from 2.2 days to 38.3 days with a mean of
8.3 days and coefficient of variation of 0.7. There are significant
differences in LBL amongst species (ANOVA on median LBL controlling for
sex and pre-release feeding: \emph{F}\textsubscript{37,194} = 2.5,
\emph{p} \textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{38}\), \emph{p}\textless{}0.01).

The estimated median LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.5, 5.0 and 6.9 days respectively with an overall estimate of 4.6 days
(Fig. 2). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

The MRR experiments included a mixture of male-only and female-only
releases, and releases of both sexes. We estimated average male and
female LBL at the genus level (Fig. 3; there were too few studies to
make comparisons at the species level). There was a consistent trend for
females to live longer than males for each of the genera, with the
difference largest for \emph{Aedes} (2.5 days; fraction of pairwise
posterior samples of females versus males where difference was less than
zero, p\textless{}0.01), followed by \emph{Anopheles} (2.0 days; p=0.17)
and \emph{Culex} (0.3 days; p=0.34). Overall, female mosquitoes were
estimated to live 0.9 days longer than males (p=0.10).

@@

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. 4). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
(\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}) individually (Fig. S\#).

@@

Dissection allows the number of completed gonotrophic cycles to be
counted and from this the mean number of cycles before death was
estimated. Overall, the mean number of cycles completed in a lifetime
was 1.2 (Fig. 5; posterior median) and across the 131 studies, 95\% of
the individual time series estimates were less than 3 (Fig. 6). The
estimated greatest number of cycles was for \emph{Anopheles sergentii}
(2.5 cycles; posterior median) which is adapted to desert conditions (it
is known as the ``oasis vector'' of malaria) and may have evolved
greater longevity. The important African malaria vector \emph{An.
gambiae s.l.} was estimated to be the second longest living (1.9 cycles;
posterior median). The smallest estimated mean number of gonotrophic
cycles was for \emph{Anopheles} \emph{bellator} (0.5 cycles; posterior
median) which transmits malaria in Brazil's Atlantic Forest. There were
significant differences in estimated lifetime gonotrophic cycles amongst
species (ANOVA: \emph{F}\textsubscript{24,106} =2.2, \emph{p}
\textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{24}\), \emph{p}\textless{}0.01).

The estimated lifetime gonotrophic cycles for the different genera were
\emph{Anopheles,} 1.4; \emph{Mansonia}, 1.1; \emph{Culex,} 1.0; and
\emph{Aedes} 0.8 (Fig. 5) and the differences between the genera were
significant (ANOVA: \emph{F}\textsubscript{3,127} =3.4, \emph{p} =0.02;
the non-parametric Kruskal Wallace: \(\chi_{3}^{2} = 21.7\),
\emph{p}\textless{}0.01).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the Methods (Fig. S\#). For ten species, we had enough data from both
species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. 7), and in only one case -- for \emph{A.
darlingi} - (Table S\#) there was a significant difference in the
time-series level LBLs.

@@

In Fig. 8 we compare the performance of the six models for describing
lifespan in MRR studies of 33 species using K-fold cross-validation. We
categorised the evidence for age-dependent mortality in each species
according to the performance of the five age-dependent models versus the
exponential: `+' indicated that all age-dependent models outperformed
the exponential; `?' indicated that the exponential outperformed one or
more age-dependent models; and `-' indicated that the exponential
performed at least as well as all other models. Overall, we estimated
that there were 8 `+' species, where age-dependent mortality fit the
data better; 11 `?' species where the evidence was mixed; and 14 species
where constant mortality models performed at least as well. The species
where age-dependent mortality best fit the data included the major
vector of dengue fever, Zika and chikungunya, \emph{Ae. Aegypti}. These
studies also tended to include multiple release MRR studies which, on
average, were conducted over a longer period of time than the others,
which may be why we failed to detect age-dependence in the latter (Fig
S\#).

In Fig. 9, we compare the performance of the six models for describing
lifespan in dissection studies of 25 species using K-fold
cross-validation, and categorise the evidence in the same way as for the
MRR analysis. By our metric, we determined that there were only two
species with evidence for age-dependent mortality (the major African
malaria vector \emph{A. gambiae s.l.} and \emph{A. minimus}, a malaria
vector in Asia).

@@

We can use the posterior parameter estimates from our Bayesian analysis
to estimate the fraction of mosquitoes that live beyond a certain age.
In order to transmit a disease, a mosquito must live longer than the
length of the intrinsic incubation period (the time taken for a pathogen
ingested in one blood meal to be ready to be transmitted during a future
feeding event). This is a lower bound as it does not include the waiting
time to find a host after feeding or egg maturation. In Fig. 10 we plot
the fraction of the mosquito population that pass this threshold using
estimates from both MRR and dissection studies for vector species (see
SOM for references used to identify species as vectors) and their most
significant diseases.

@@

\emph{Ae. aegypti} and \emph{Ae. albopictus} are the main vectors of
dengue, chikungunya and Zika viruses. Because of their short intrinsic
incubation periods a greater fraction of mosquito potentially live long
enough to transmit diseases (Fig. 10), rising to a maximum of 84\% for
\emph{Ae. albopictus} transmitting chikungunya.

@@

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. 7).
In is reassuring that there is correlation between estimates from both
approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with a posterior median LBL of 8.7 days (median estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 10.0 days (median
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara (Sinka
et al., 2010), where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.4 days
(median estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 7.6 days (median estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in
estimates was for \emph{Culex} (a posterior median of 2.5 days from the
MRR versus 4.4 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (4.8 versus 5.0 days). By contrast for
\emph{Aedes} the estimates from the MRR studies (6.9 days) exceed those
of dissection-based studies (3.5 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 4.6 days versus 5.1
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

@@

In this work, we have used modern statistical methods to synthesise
precious field data conducted by entomologists past and present, to
produce lower bound estimates of mosquito lifespan. The importance of
vector mortality for disease transmission has long been recognised,
however, since even before 1957, when George Macdonald formulated the
now famous Ross-Macdonald equation of R\textsubscript{0} for malaria.
Indeed, the recent declines in malaria prevalence in Sub-Saharan Africa
were likely due to the upscaling of interventions (insecticide-treated
bednets and indoor residual spraying) that aim to reduce mosquito
lifespan (Bhatt et al., 2015). Worryingly, resistance to pyrethroids,
the only class of insecticide used in current insecticide-treated
bednets and likely the only product to come to market in the near
future, has been determined to be widespread and increasing in intensity
across Sub-Saharan Africa (WHO reference 2018). This alarming trend
highlights the need for continued MRR and dissection-based studies to
monitor the effectiveness of bednets and determine whether more
expensive alternatives, such as nets incorporating piperonyl butoxide be
deployed. It also emphasises the need for investment in new tools for
real time monitoring of mosquito populations. In recent years, there has
been considerable funding allocated to molecular and genomic research
into mosquitoes that strengthens existing interventions and suggest
novel control strategies. Without commensurate funding allocated to
applied vector ecology, our lack of knowledge in this area threatens our
opportunity to capitalise on molecular advances and potentially hinders
our ability to control of mosquito-borne disease.

>>>>

\usepackage{xr}
\externaldocument{figures}

@@

MRR estimates the length of time a mosquito remains alive and is still
in the area available for recapture. It thus provides a lower bound to
lifespan which we shall refer to as LBL. In 187 of the 230 MRR time
series the estimated LBL was less than 10 days (Fig. \ref{fig:mrr_lifespan_individualEstimates}). The smallest
estimate was \textless{}1 day for the Asian malaria vector
\emph{Anopheles subpictus s.l.} which is unfeasibly short and almost
certainly reflects dispersal out of the recapture zone or a violation of
the assumptions of our analyses. The longest estimate was 18.3 days for
the temperate species \emph{Aedes simpsoni s.l.} which is a vector of
yellow fever in Africa. There are multiple data sets for the most
important vector species such as \emph{Anopheles gambiae, Aedes aegypti}
and \emph{albopictus} and \emph{Culex tarsalis} all of which show
considerable variation. For example, there are 54 estimates of LBL for
\emph{A. aegypti} which range from 2.2 days to 38.3 days with a mean of
8.3 days and coefficient of variation of 0.7. There are significant
differences in LBL amongst species (ANOVA on median LBL controlling for
sex and pre-release feeding: \emph{F}\textsubscript{37,194} = 2.5,
\emph{p} \textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{38}\), \emph{p}\textless{}0.01).

The estimated median LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.5, 5.0 and 6.9 days respectively with an overall estimate of 4.6 days
(Fig. \ref{fig:mrr_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

The MRR experiments included a mixture of male-only and female-only
releases, and releases of both sexes. We estimated average male and
female LBL at the genus level (Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}; there were too few studies to
make comparisons at the species level). There was a consistent trend for
females to live longer than males for each of the genera, with the
difference largest for \emph{Aedes} (2.5 days; fraction of pairwise
posterior samples of females versus males where difference was less than
zero, p\textless{}0.01), followed by \emph{Anopheles} (2.0 days; p=0.17)
and \emph{Culex} (0.3 days; p=0.34). Overall, female mosquitoes were
estimated to live 0.9 days longer than males (p=0.10).

@@

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
(\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}) individually (Fig. S\#).

@@

Dissection allows the number of completed gonotrophic cycles to be
counted and from this the mean number of cycles before death was
estimated. Overall, the mean number of cycles completed in a lifetime
was 1.2 (Fig. \ref{fig:dissection_lifetimes_exponential}; posterior median) and across the 131 studies, 95\% of
the individual time series estimates were less than 3 (Fig. \ref{fig:dissection_individualEstimates_allSpecies}). The
estimated greatest number of cycles was for \emph{Anopheles sergentii}
(2.5 cycles; posterior median) which is adapted to desert conditions (it
is known as the ``oasis vector'' of malaria) and may have evolved
greater longevity. The important African malaria vector \emph{An.
gambiae s.l.} was estimated to be the second longest living (1.9 cycles;
posterior median). The smallest estimated mean number of gonotrophic
cycles was for \emph{Anopheles} \emph{bellator} (0.5 cycles; posterior
median) which transmits malaria in Brazil's Atlantic Forest. There were
significant differences in estimated lifetime gonotrophic cycles amongst
species (ANOVA: \emph{F}\textsubscript{24,106} =2.2, \emph{p}
\textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{24}\), \emph{p}\textless{}0.01).

The estimated lifetime gonotrophic cycles for the different genera were
\emph{Anopheles,} 1.4; \emph{Mansonia}, 1.1; \emph{Culex,} 1.0; and
\emph{Aedes} 0.8 (Fig. \ref{fig:dissection_lifetimes_exponential}) and the differences between the genera were
significant (ANOVA: \emph{F}\textsubscript{3,127} =3.4, \emph{p} =0.02;
the non-parametric Kruskal Wallace: \(\chi_{3}^{2} = 21.7\),
\emph{p}\textless{}0.01).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the Methods (Fig. S\#). For ten species, we had enough data from both
species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - (Table S\#) there was a significant difference in the
time-series level LBLs.

@@

In Fig. \ref{fig:mrr_elpd}, we compare the performance of the six models for describing
lifespan in MRR studies of 33 species using K-fold cross-validation. We
categorised the evidence for age-dependent mortality in each species
according to the performance of the five age-dependent models versus the
exponential: `+' indicated that all age-dependent models outperformed
the exponential; `?' indicated that the exponential outperformed one or
more age-dependent models; and `-' indicated that the exponential
performed at least as well as all other models. Overall, we estimated
that there were 8 `+' species, where age-dependent mortality fit the
data better; 11 `?' species where the evidence was mixed; and 14 species
where constant mortality models performed at least as well. The species
where age-dependent mortality best fit the data included the major
vector of dengue fever, Zika and chikungunya, \emph{Ae. Aegypti}. These
studies also tended to include multiple release MRR studies which, on
average, were conducted over a longer period of time than the others,
which may be why we failed to detect age-dependence in the latter (Fig
S\#).

In Fig. \ref{fig:dissection_elpd}, we compare the performance of the six models for describing
lifespan in dissection studies of 25 species using K-fold
cross-validation, and categorise the evidence in the same way as for the
MRR analysis. By our metric, we determined that there were only two
species with evidence for age-dependent mortality (the major African
malaria vector \emph{A. gambiae s.l.} and \emph{A. minimus}, a malaria
vector in Asia).

@@

We can use the posterior parameter estimates from our Bayesian analysis
to estimate the fraction of mosquitoes that live beyond a certain age.
In order to transmit a disease, a mosquito must live longer than the
length of the intrinsic incubation period (the time taken for a pathogen
ingested in one blood meal to be ready to be transmitted during a future
feeding event). This is a lower bound as it does not include the waiting
time to find a host after feeding or egg maturation. In Fig. \ref{fig:eip}, we plot
the fraction of the mosquito population that pass this threshold using
estimates from both MRR and dissection studies for vector species (see
SOM for references used to identify species as vectors) and their most
significant diseases.

@@

\emph{Ae. aegypti} and \emph{Ae. albopictus} are the main vectors of
dengue, chikungunya and Zika viruses. Because of their short intrinsic
incubation periods a greater fraction of mosquito potentially live long
enough to transmit diseases (Fig. \ref{fig:eip}), rising to a maximum of 84\% for
\emph{Ae. albopictus} transmitting chikungunya.

@@

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with a posterior median LBL of 8.7 days (median estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 10.0 days (median
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara (Sinka
et al., 2010), where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.4 days
(median estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 7.6 days (median estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in
estimates was for \emph{Culex} (a posterior median of 2.5 days from the
MRR versus 4.4 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (4.8 versus 5.0 days). By contrast for
\emph{Aedes} the estimates from the MRR studies (6.9 days) exceed those
of dissection-based studies (3.5 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 4.6 days versus 5.1
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

@@

The importance of vector mortality for disease transmission has long been recognised,
since even before 1957, when George Macdonald formulated the
now famous Ross-Macdonald equation of R\textsubscript{0} for malaria. Indeed, the recent declines in malaria prevalence in Sub-Saharan Africa were likely due to upscaling of interventions (insecticide-treated bednets and indoor residual spraying) that aim to reduce mosquito lifespan (Bhatt et al., 2015). Worryingly, resistance to pyrethroids, the only class of insecticide used in current insecticide-treated bednets and likely the only product to come to market in the near
future, has been determined to be widespread and increasing in intensity
across Sub-Saharan Africa (WHO reference 2018). This alarming trend
highlights the need for continued MRR and dissection-based studies to
monitor the effectiveness of bednets and determine whether more
expensive alternatives, such as nets incorporating piperonyl butoxide be
deployed. It also emphasises the need for investment in new tools for
real time monitoring of mosquito populations. In this work, we have used modern statistical methods to synthesise precious field data conducted by entomologists past and present, to produce lower bound estimates of mosquito lifespan. In recent years, there has been considerable funding allocated to molecular and genomic research
into mosquitoes that strengthens existing interventions and suggest
novel control strategies. Without commensurate funding allocated to
applied vector ecology, our lack of knowledge in this area threatens our
opportunity to capitalise on molecular advances and potentially hinders
our ability to control of mosquito-borne disease.
